Replacing and additive horizontal \\ gene transfer in {\em Streptococcus}

### 2011-12-13 doc/manuscript.tex
A supplementary is added.

% TODO:
% Figure S\label{fig:calling-transfers} was not referred in the text.(Delete this?) 
% Figure S\label{fig:mowgli-sim} was not referred in the text. (Write comments)
% Table S\label{tab:sim-three} was not referred in the text. (Delete this?)
% 
%
% * Finalize section: Distribution of transfer events along the genome
% * Finalize section: Supp. Materials.
% Supp. Methods of Matt's modifications of Mowgli's method
% 
% Backup of the work
% swiftgen:/usr/projects/strep/gainloss/recombination

@@

\usepackage{times}

@@

% FIXME: Remove this package when submitting the manuscript.
\usepackage{url}

@@

% Text layout
\oddsidemargin 0in
\evensidemargin 0in
\topmargin -.5in
\textwidth 6.5in
\textheight 9in
%\topmargin 0.0cm
%\oddsidemargin 0.5cm
%\evensidemargin 0.5cm
%\textwidth 16cm 
%\textheight 21cm

@@

% \@ifundefined{showcaptionsetup}{}{%
% \PassOptionsToPackage{caption=false}{subfig}}
% \usepackage{subfig}

@@

% Leave date blank
\date{}

@@

The prominent role of horizontal gene transfer (HGT) in the evolution of
bacteria is now well documented, yet few studies have differentiated
between the mechanisms of homologous recombination and additive integration.
%as sources of HGT.  
These processes leave distinct phylogenetic signatures, because homologous
recombination causes a gene in one lineage to be replaced by a homolog from
another lineage (``replacing HGT''), while additive integration causes
addition without replacement (``additive HGT'').  Here we make use of these
signatures in a genome-wide investigation of HGT in the important human
pathogen {\em Streptococcus pyogenes} (SPY) and its close relatives {\em
  S.\ dysgalactiae} subspecies {\em equisimilis} (SDE) and {\em S.\
  dysgalactiae} subspecies {\em dysgalactiae} (SDD).  Using recently
developed statistical models and computational methods, we find evidence
for abundant gene flow of both kinds within the SPY and SDE clades, and of
generally reduced levels of exchange between SPY/SDE and SDD.  In addition,
our analysis strongly supports a previously reported, but unconfirmed,
finding of a pronounced asymmetry in SPY-SDE gene flow, favoring the
SPY-to-SDE direction, which may be associated with increasing virulence of
pathogenic SDE.  We find that this enrichment is driven by replacing
transfers, and likely reflects frequent homologous recombination between
co-occurring SPY and SDE cells in human hosts.  We find that virulence
genes are correlated with transfer events, but this correlation is driven
by additive, not replacement, HGTs.  The genes affected by additive HGTs
are enriched for transposition-related functions, while replacement HGTs
appear to influence a more diverse set of genes.  These findings shed new
light on the manner in which HGT has shaped pathogenic bacterial genomes
during relatively recent evolutionary time.

@@

During the last twenty years, it has become increasingly apparent that
horizontal gene transfer (HGT) has played a major role in genomic
evolution, particularly among bacteria and other microbes
\citep{Smith1992,Smith1993,Ochman2000,Koonin2001}.  Indeed, it is likely
that HGT has been sufficiently prevalent throughout evolutionary history
that no present-day gene can trace an unbroken history of vertical descent
to a common ancestor for all species \citep{Zhaxybayeva2011}.  This would
mean that no single phylogenetic marker is a reliable indicator of the tree
of life.

@@

In this article, we take a fresh look at the issue of HGT in the {\em
  Pyogenes} group, making use of newly available complete genome sequences
for SPY, SDE, and the related species {\em S.\ dysgalactiae} ssp.\ {\em
  dysgalactiae} (SDD), and new statistical and parsimony-based methods for
analysis.  The combined use of these methods allows us to examine both
replacing and additive HGTs, and compare their genome-wide effects.  We
find evidence for abundant gene flow within SPY, within SDE, and between
SPY and SDE, and for greatly reduced exchange between SPY/SDE and SDD.  In
addition, our genome-wide analysis supports Kalia et al.'s
\citeyearpar{Kalia2001} previously unconfirmed finding of a pronounced
preference for the SPY-to-SDE direction in SPY-SDE gene flow.
Interestingly, this property appears to hold for replacing but not for
additive transfers.  More generally, we find evidence for an association
between virulence genes and additive, but not replacing, HGTs.  Our results
have been made available as tracks in our recently released {\em
  Streptococcus} genome browser.

@@

Our primary analysis was based on five complete genome sequences, including
two representatives of {\em S.\ pyogenes} (accessions NC\_004070
[\citealp{Beres2002}; SPY1] and NC\_008024 [\citealp{Beres2006}, SPY2]),
two of {\em S.\ dysgalactiae} ssp.\ {\em equisimilis} (CP002215
[\citealp{Suzuki2011}; SDE1] and NC\_012891 [\citealp{Shimomura2011};
SDE2]), and one of {\em S.\ dysgalactiae} ssp.\ {\em dysgalactiae}
(AEGO00000000 [\citealp{Suzuki2011}; SDD]).  In addition, we used 
\textit{S.\ equi} ssp.\textit{\ equi} strain 4047 (NC\_012471
[\citealp{Holden2009}; SEE]) as an outgroup in our parsimony-based
analysis (supplementary table S\ref{tab:genome}).

For the model-based analysis of replacing transfers, we obtained
high-quality alignments of the five genomes (SPY1 \& 2, SDE1 \& 2, and
SDD) using the 
pipeline detailed by Didelot and colleagues
\citep{Didelot2007,Didelot2010}.  Briefly, we first aligned the genomes
using progressiveMauve v2.3.1 \citep{Darling2004,Darling2010} with default
options (supplementary fig.\ S\ref{fig:mauve}), and then used stripSubsetLCBs (from the
ClonalOrigin package; \citealp{Didelot2010}) to identify 276 blocks of
sequence alignments that exceed a length threshold of 1500 bp.  This
conservative threshold was selected to avoid biases from boundary effects
in short alignment blocks, based on a preliminary analysis.  Two blocks
containing long gaps were excluded from the analysis.
The final set of 274
alignment blocks was 1147632 nucleotide long.
Of 1951 genes of SPY1 genome 1125 genes were covered by the blocks.
Note that this alignment
procedure effectively focused on the ``core genome,'' by discarding regions
that were not conserved among the five species in question.  The 1125 genes
contained 0.17\% of gaps on average.

For the parsimony-based analysis, we began with 2314 gene family
clusters representing the six genomes (including the SEE outgroup)
from \cite{Suzuki2011}.  We aligned protein sequences corresponding to
each family using MUSCLE \citep{Edgar2004a}, and then obtained
nucleotide alignments by reverse translation, using the genomic
sequences as a guide.  Next we identified likely intragenic
recombinations using the Single Breakpoint Recombination (SBP) method
from the HyPhy package \citep{KosakovskyPond2006}.  Based on the
Akaike Information Criterion (AIC), SBP identified at least one
topology-altering recombination in 47.5\% (1094) of gene
families. These alignments were split into putatively non-recombining
gene fragments.  The final set of alignments contained 3408 gene
fragment families, 1794 of which contained one gene in each of the six
species.  Families ranged as small as two genes (417 fragments) to as
large as 59 (2 fragments) (supplementary fig.\ S\ref{fig:famsizes}).  Notably, in
contrast to the genome-wide alignments produced by progressiveMauve,
these alignments were representative of the ``pan genome,'' including
regions that frequently turn over, rather than just the ``core
genome'' \citep[e.g.,][]{Tettelin2005,Lefebure2007}.

@@

{\bf ClonalFrame and ClonalOrigin.}  To study replacing HGTs, we made use
of the coalescent-based model of homologous recombination and the Markov
chain Monte Carlo (MCMC) algorithm recently developed by
\cite{Didelot2010}.  The method was implemented in a 
program, called ClonalOrigin, that could produce samples from the posterior
distribution of {\em recombinant trees} given a set of alignment blocks. A
recombinant tree consists of a rooted tree with branch length, called a clonal
frame, for the five bacterial genomes,
which is estimated in a pre-processing step, augmented by a set of
{\em recombinant edges}. Each recombinant edge is defined by the two points
along the branches of clonal frame corresponding to the {\em donor} and
{\em recipient} in the hypothesized recombination event, and by the start
and end of the genomic interval that is affected by the event.  The donor
must pre-date the recipient, and a delayed coalescence is possible
(see \textbf{Supplementary Material} for details of summarizing recombinant
trees and edges). 
Another program, called ClonalFrame
\citep{Didelot2007}, can be used to infer the phylogeny that serves as the
basis of the recombinant trees.  ClonalFrame makes use of a similar, but
somewhat simpler probabilistic model, which permits inference
of the clonal frame. % and a set of model parameters.

{\bf Fitting the model.}  We used
ClonalFrame (v1.1) to estimate 
a clonal frame for the five-way
alignments, running the algorithm for $10^4$ burn-in iterations followed by
another $10^4$ sampling iterations.  A series of seven additional sampling
runs with 
random initializations converged to an essentially identical clonal frame, % model parameters,
indicating stable convergence.  We then used an initial run of
ClonalOrigin (subversion r19) to estimate global parameters including
the mutation rate, recombination rate, and recombinant tract length. We
applied the method separately to each of the 274 blocks, in all cases using
$10^6$ burn-in iterations, $10^7$ sampling iterations, and sub-sampling
every $10^5$ iterations. A global estimate for each parameter was then
obtained by taking the block-length weighted median of the block-specific
posterior means, which down-weighed less reliable estimates obtained from
shorter alignment blocks.  We generally followed \cite{Didelot2010} in 
fitting the model. % , with some minor differences.

{\bf Sampling recombinant trees}.  Next we ran a second stage of ClonalOrigin
MCMC
on all of the alignment blocks by fixing the clonal frame and global model
parameters at their pre-estimated values, and sampled
recombinant trees.  For each block, we ran the sampler for $10^7$ burn-in
iterations and $10^8$ sampling iterations, sub-sampling every $10^5$
iterations, resulting in 1001 recombinant trees per alignment block.  We
performed a replicate of the two stages of the ClonalOrigin inference to
ensure adequate convergence of the sampler.

% A few alignment blocks were not successfully finished even after a month of
% computation. Because those were not expected to affect the global estimates of
% population parameters significantly, we used only finished blocks to estimate
% mutation rate, recombination rate, and recombinant tract length.

{\bf Summary statistics.}  We used various summary statistics to describe
the sampled recombinant trees (see \textbf{Supplementary Material} for details). 
First, we considered the distribution of
tree topologies at each site in the alignments, by computing the site-wise
relative frequencies of all 105 possible rooted tree topologies relating five
taxa. Second, we measured frequencies in the posterior samples
of all possible types of recombinant edge, where an edge
type is defined by the branches of the clonal frame associated with its
donor and recipient.
%also searched the posterior sample for transfer scenarios that were
%significantly enriched, compared to the prior expectation defined by the
%clonal frame and the global estimates of the three population
%parameters. We followed the approach of \cite{Didelot2010} and defined
%transfer types by the branches of the clonal frame associated with each
%recombinant edge. 
%For each transfer type, we recorded the average number of
%recombinant edges of that type sampled in the posterior sample of
%recombinant trees, 
We compared these posterior frequencies with 
expected values under the prior distribution assumed by the model.
%as well as the expected number of such edges implied by
%the prior distribution assumed by the model. 
%These counts were then used to
%find types of transfer that showed a high deviation in our inference from
%their expected counts. 
To assess significance of the ratio of posterior to prior frequencies of
recombinant edges we performed a simulation; 
this empirical approach also 
had the advantage of correcting for possible biases in the sampling
process. 
\textbf{CHECK:}
We simulated the model to generate 100 data sets, for which we proceeded with
the inference of the ratio of posterior to prior frequencies of recombinant
edges. Comparing the obtained null distribution of the ratios from the
simulation and the ratio from the real data analysis, we assessed the
significance of the ratio.
% (GIVE VERY BRIEF SUMMARY OF SIMULATION APPROACH)
%Ilan: point to supplementary methods for the simulation tests?
Third,
we traced the posterior
probabilities of transfer types in a position-specific manner. 
%For each
%distinct type of transfer (determined by a pair of donor and recipient
%edges), and each site in the core genome alignment, we computed the
%posterior probability that that site had been affected by a transfer of
%that type. 
Finally, we defined a general, branch-independent
 {\em recombination intensity} at each site by 
summing the posterior probabilities of recombination.
In some cases, we also computed intensities for particular types of
replacing transfers, such as 
those between the SPY and SDE clades, or those producing topologies
different from the clonal frame, or \textit{topology-altering} transfers.  
% CHECK: Justify why we used only topology-altering edges for computing
% recombination intensity.
Note that \cite{Didelot2010} described the second and third summarization
procedures.

@@

{\bf Mowgli.} To study additive HGTs, we used a parsimony-based method
for reconciling gene and species trees called Mowgli \citep{Doyon2011}.
Given a gene tree and a species tree, Mowgli finds a reconciliation scenario 
that minimizes the number of gene duplication, loss, and
transfer (DLT) events by considering tree topologies with branching orders, also
known as labeled histories \citep{Edwards1970}.  We estimated
unrooted gene trees for the nucleotide alignments for each of our
putatively non-recombining gene fragment using RAxML \citep{Stamatakis2006},
then applied Mowgli to each of these trees together with the species
tree estimated by ClonalFrame (augmented with the SEE outgroup).
Since Mowgli required rooted gene trees, we ran Mowgli for all possible
rootings of each gene tree, and chose the scenario that produced the
minimum DLT score. In the case of multiple most parsimonious
reconciliations, one was chosen uniformly at random.

{\bf Distinguishing between replacing and additive HGTs.}  The transfer
events inferred by Mowgli had to be classified as replacing or
additive in a post-processing step because the program did not
distinguish between these types of events.  A naive approach 
would be to infer a replacing HGT whenever an inferred transfer
coincided exactly with an inferred loss in the recipient lineage.  However,
loss events were difficult to place accurately because they were frequently
pushed toward the root of the tree due to reconstruction error or parallel
losses in multiple lineages.  Therefore, we classified a transfer event as
additive whenever a descendant of the recipient lineage in the species tree
contained both a gene descended from the transferred copy and a gene not
descended from that copy.  All other transfers were considered replacing
HGTs.  This was conservative about calling additive HGTs, and would
label any HGT that could plausibly be explained by replacing transfers
as such (see fig.\ \ref{fig:hgt}).

{\bf Statistical enrichment}. As with replacing
HGTs, we classified transfer events by donor and recipient branch, 
%sought to identify over-represented scenarios of additive HGT. 
%and identified distinct types of additive
%transfers by their donor and recipient branches in the species
%tree, 
and recorded the number of inferred additive HGTs of each type.
%as well as the prior expected number of transfers
%of that type. 
We compared these numbers to expectations based on simulations that
assumed 
constant
rates of duplication, loss, and transfer across the species tree, given 
the total numbers of events inferred across all 2314 gene families
divided by the total branch lengths of the trees
(see \textbf{Supplementary Material}).

@@

We performed a series of analyses to search for certain classes of genes
enriched for replacing or additive of gene transfers. In this analysis,
each gene, or gene family, was associated with a certain score. For
replacing gene transfers we used the recombination intensity along
a given gene, and for additive gene transfers we used the number of
additive transfers inferred for a given family. Mann-Whitney tests were
used to identify sets of genes that showed significantly elevated values in
each score.  A 5\% False Discovery Rate (FDR) correction
\citep{Benjamini1995} was used to determine an appropriate significance
threshold for our p-values. We applied the association test for gene
classes defined by Gene Ontology (GO) terms assigned  
%using the method described in Suzuki et al.\ \cite{Suzuki2011}. In short, 
by comparing the Streptococcus genes to bacterial proteins from the
Uniref90 database using blastP, and then assigning the same GO
classification as the target gene of the uniProt GOA database, if the
match had an E-values less than $1.0\times10^{-5}$. We also classified a
category of \textit{virulence genes} using the virulence profile obtained
by \cite{Suzuki2011}. In both cases, a gene family was
associsated with a certain classification if any of its genes were
associated with that classification.

@@

\subsection*{Clonal frame and global parameter estimates}
% The 5-species tree was
% (((1:0.045557,2:0.045557)6:0.195330,3:0.240887)8:0.089874,(4:0.074544,5:0.074544)7:0.256218)9:0.000000
By applying ClonalFrame to our five-way genomic alignments 
(supplementary fig.\ S\ref{fig:mauve}), we obtained a
phylogeny in which the pairs of SPY and SDE samples were grouped together,
as expected, and in which SDD and SDE formed a clade, with SPY as an
outgroup (fig.\ \ref{fig:tree5}).  This tree was consistent with previous
results from 16S rRNA sequences \citep{Facklam2002} and with genome-wide
trends \citep{Suzuki2011}.  Despite the grouping of SDD and SDE, their
estimated divergence was substantial, at $\sim$0.48 substitutions per site.  
The estimated divergence between SPY and SDD/SDE was $\sim$0.66
substitutions per site.  
%These estimates correspond to average
%genomic divergence times of roughly 660 and 900 thousand years,
%respectively, assuming a mutation rate of $5 \times 10^{-10}$ substitutions
%per site per generation \citep{Ochman2003} and two cell divisions per day
%\citep{XX}.  
An initial analysis with ClonalOrigin (see \textbf{Methods}) produced an
estimated
per-site population-scaled mutation rate of $\theta = $ 0.081
[interquartile range across blocks: (0.067, 0.094)], a per-site
population-scaled recombination rate of $\rho = $ 0.012 (0.006, 0.019), and
an average recombinant tract length of $\delta = $ 744 (346, 2848) bp.  The
full distributions of these quantities across alignment blocks are shown in
supplementary
figure S\ref{fig:scatter3}.  Our estimates implied a ratio of $\rho/\theta
= 0.15$.  For comparison, \cite{Didelot2010} obtained estimates of $\theta
= 0.044$, $\rho = 0.017$, $\delta = 236$, and $\rho/\theta = 0.405$ for the
{\em Bacillus cereus} group, but differences of this magnitude were not
surprising, given likely differences in effective population sizes,
recombination rates, and mutation rates \citep[e.g.,][]{Feil2001}.  We
found that estimates of the recombination tract length, $\delta$, were
quite sensitive to our threshold for minimum block length, with the
inclusion of short alignments producing much larger estimates due to edge
effects.  We experimented with a range of thresholds and selected one (1500
bp) at which the estimates stabilized.

@@

% accurate to say "recent"?  has been going on for some time
There has been a great deal of recent interest in the process of horizontal
gene transfer (HGT) and the manner in which it has influenced phylogenetic
relationships, particularly among bacteria \citep{}.  Most previous studies,
however, have either simply examined patterns of phylogenetic discordance,
without differentiating between additive and replacing HGTs \citep{}, or
have focused specifically on the process of homologous recombination,
corresponding to replacing HGTs
\citep{Didelot2007,Didelot2010}.
% too strong?  neither are we
% cite Feil?   Marri.  LEfebre.  recent review (Doolittle?)
% another homologous recombination review -- see Awadalla
To our knowledge, this is the first study to examine both
processes on a genome-wide scale, using both model- and parsimony-based
methods.

>>>>

% TODO:
% 1. Not yet referenced figure - \label{fig:mowgli-sim}
% 2. Not yet referenced figure - \label{fig:sim2}
% 3. Not yet referenced figure - \label{fig:sim3}
% 
% Backup of the work
% swiftgen:/usr/projects/strep/gainloss/recombination

@@

\usepackage{times}
% \usepackage[T1]{fontenc}
% \usepackage{ae,aecompl}

@@

@@

\usepackage{xr}
\externaldocument{supplementary}

% Text layout
\oddsidemargin 0in
\evensidemargin 0in
\topmargin -.5in
\textwidth 6.5in
\textheight 9in

@@

@@

@@

The prominent role of horizontal gene transfer (HGT) in the evolution of
bacteria is now well documented, yet few studies have differentiated
between the mechanisms of homologous recombination and additive
integration.  These processes leave distinct phylogenetic signatures,
because homologous recombination causes a gene in one lineage to be
replaced by a homolog from another lineage (``replacing HGT''), while
additive integration causes addition without replacement (``additive
HGT'').  Here we make use of these signatures in a genome-wide
investigation of HGT in the important human pathogen {\em Streptococcus
  pyogenes} (SPY) and its close relatives {\em S.\ dysgalactiae} subspecies
{\em equisimilis} (SDE) and {\em S.\ dysgalactiae} subspecies {\em
  dysgalactiae} (SDD).  Using recently developed statistical models and
computational methods, we find evidence for abundant gene flow of both
kinds within the SPY and SDE clades, and of reduced levels of exchange
between SPY and SDD.  In addition, our analysis strongly supports a
previously reported, but unconfirmed, finding of a pronounced asymmetry in
SPY-SDE gene flow, favoring the SPY-to-SDE direction, which may be
associated with increasing virulence of pathogenic SDE.  We find much
stronger evidence for SPY-SDE gene flow among replacing rather than
additive transfers, suggesting a primary influence from homologous
recombination between co-occurring SPY and SDE cells in human hosts.
Virulence genes are correlated with transfer events, but this correlation
is driven by additive, not replacing, HGTs.  The genes affected by additive
HGTs are enriched for functions having to do with transposition,
recombination, and DNA integration, while replacing HGTs appear to
influence a more diverse set of genes.  Additive transfers are also
associated with evidence of positive selection.  These findings shed new
light on the manner in which HGT has shaped pathogenic bacterial genomes
during relatively recent evolutionary time.

@@

During the last twenty years, it has become increasingly apparent that
horizontal gene transfer (HGT) has played a major role in genomic
evolution, particularly among bacteria and other microbes
\citep{Smith1992,Smith1993,Ochman2000,Koonin2001}.  Indeed, it is likely
that HGT has been sufficiently prevalent throughout evolutionary history
that no present-day gene can trace an unbroken history of vertical descent
to a common ancestor for all species \citep{Zhaxybayeva2011}.  This would
mean that no phylogenetic marker can be used to reconstruct a
universal tree of life.

@@

In this article, we take a fresh look at the issue of HGT in the {\em
  Pyogenes} group, making use of newly available complete genome sequences
for SPY, SDE, and the related species {\em S.\ dysgalactiae} ssp.\ {\em
  dysgalactiae} (SDD), as well as new statistical and parsimony-based
methods for 
analysis.  The combined use of these methods allows us to examine both
replacing and additive HGTs, and compare their genome-wide effects.  We
find evidence for abundant gene flow within SPY, within SDE, and between
SPY and SDE, and for greatly reduced exchange between SPY and SDD.  In
addition, our genome-wide analysis supports Kalia et al.'s
\citeyearpar{Kalia2001} previously unconfirmed finding of a pronounced
preference for the SPY-to-SDE direction in SPY-SDE gene flow.
Interestingly, this property is much more evident for replacing than for
additive transfers.  We also examine correlations of gene transfer
events with functional categories and positive selection.
%find evidence for an association
%between virulence genes and additive, but not replacing, HGTs.  
Our results
have been made available as tracks in our recently released {\em
  Streptococcus} genome browser.

@@

Our primary analysis was based on five complete genome sequences, including
two representatives of {\em S.\ pyogenes} (accessions NC\_004070
[\citealp{Beres2002}; SPY1] and NC\_008024 [\citealp{Beres2006}, SPY2]),
two of {\em S.\ dysgalactiae} ssp.\ {\em equisimilis} (CP002215
[\citealp{Suzuki2011}; SDE1] and NC\_012891 [\citealp{Shimomura2011};
SDE2]), and one of {\em S.\ dysgalactiae} ssp.\ {\em dysgalactiae}
(CM001076 [\citealp{Suzuki2011}; SDD]).  In addition, we used 
\textit{S.\ equi} ssp.\textit{\ equi} strain 4047 (NC\_012471
[\citealp{Holden2009}; SEE]) as an outgroup in our parsimony-based
analysis (supplementary table \ref{tab:genome}).

For the model-based analysis of replacing transfers, we obtained
high-quality alignments of SPY1, SPY2, SDE1, SDE2, and SDD using the
pipeline detailed by Didelot and colleagues
\citep{Didelot2007,Didelot2010}.  Briefly, we first aligned these five
genomes using progressiveMauve v2.3.1 \citep{Darling2004,Darling2010} with
default options (supplementary fig.\ \ref{fig:mauve}), and then used
stripSubsetLCBs (from the ClonalOrigin package; \citealp{Didelot2010}) to
identify 276 blocks of sequence alignments that exceed a length threshold
of 1,500 bp.  This conservative threshold was selected to avoid biases from
boundary effects in short alignment blocks, based on a preliminary
analysis.  Two blocks containing long gaps were excluded from the analysis.
The final set of 274 alignment blocks contained 1,155,016 columns with
relatively few gaps (0.2\% of characters) and covered 1,106 of the 1,951
genes in the SPY1 genome.  Note that this alignment procedure effectively
focused on the ``core genome,'' by discarding regions that were not
conserved among the five bacterial individuals in question.

For the parsimony-based analysis, we began with 2,314 gene family clusters
representing all six genomes (including SEE), as described by
\cite{Suzuki2011}.  We aligned protein sequences corresponding to each
family using MUSCLE \citep{Edgar2004a}, and then obtained nucleotide
alignments by reverse translation, using the genomic sequences as a guide.
Next we identified likely intragenic recombinations using the Single
Breakpoint Recombination (SBP) method from the HyPhy package
\citep{KosakovskyPond2006}.  Based on the Akaike Information Criterion
(AIC), SBP identified at least one topology-altering recombination in
47.5\% (1094) of gene families. These alignments were split into putative
non-recombining gene fragments.  The final set of alignments consisted of
3,408 gene fragment families, about half (1,794) of which contained
single gene from each of the six species.  The remaining families ranged
from as few as two genes (417 fragments) to as many as 59 (2 fragments)
(supplementary fig.\ \ref{fig:famsizes}).  These alignments were
representative of the ``pan genome,'' including regions that frequently
turn over (``dispensable'' regions), rather than just the ``core genome''
\citep[e.g.,][]{Tettelin2005,Lefebure2007}.

@@

{\bf ClonalFrame and ClonalOrigin.}  To study replacing HGTs, we made use
of the ClonalOrigin program recently developed by \cite{Didelot2010}.
Given a set of alignment blocks, ClonalOrigin samples from the posterior
distribution of {\em recombinant trees} using a Markov chain Monte Carlo
(MCMC) algorithm.  A recombinant tree consists of a pre-estimated rooted
phylogeny with branch lengths (a ``clonal frame''), augmented by a set of
{\em recombinant edges}. Each recombinant edge is defined by the two points
along the branches of the clonal frame that correspond to the {\em donor}
and {\em recipient} in the hypothesized recombination event, and by the
start and end of the affected genomic interval.  The donor can pre-date the
recipient, to allow for a delayed coalescence, but it cannot post-date it
(supplementary fig.\ \ref{fig:clonalorigin}).
Another program, called ClonalFrame
\citep{Didelot2007}, can be used to infer the phylogeny that serves as the
basis of the recombinant trees.  ClonalFrame makes use of a similar, but
somewhat simpler, probabilistic model.

{\bf Fitting the model.}  Following \cite{Didelot2010}, we used ClonalFrame
(v1.1) to estimate a clonal frame for our five-way alignments, running the
algorithm for $10^4$ burn-in iterations followed by another $10^4$ sampling
iterations.  A series of seven additional sampling runs with random
initializations converged to essentially identical estimates, indicating
stable convergence.  We then used an initial run of ClonalOrigin
(subversion r19) to estimate global parameters including the mutation rate,
recombination rate, and average recombinant tract length. We applied the method
separately to each of the 274 blocks, in all cases using $10^6$ burn-in
iterations, $10^7$ sampling iterations, and sub-sampling every $10^5$
iterations. A global estimate for each parameter was then obtained by
taking the block-length weighted median of the block-specific posterior
means, which down-weighed less reliable estimates obtained from shorter
alignment blocks.

{\bf Sampling recombinant trees}.  We ran ClonalOrigin a second time on all
alignment blocks, this time fixing the clonal frame and global model
parameters at their pre-estimated values, and collected samples of
recombinant trees across the genome.  For each block, we ran the sampler
for $10^7$ burn-in iterations and $10^8$ sampling iterations, sub-sampling
every $10^5$ iterations, resulting in 1,001 recombinant trees per alignment
block.  We performed a replicate of the two stages of the ClonalOrigin
inference to ensure adequate convergence of the sampler.

% A few alignment blocks were not successfully finished even after a month of
% computation. Because those were not expected to affect the global estimates of
% population parameters significantly, we used only finished blocks to estimate
% mutation rate, recombination rate, and average tract length.

{\bf Summary statistics.}  We used various summary statistics to describe
the sampled recombinant trees.
First, we recorded the relative frequencies of the 105 possible
rooted trees at each site in the alignments.  Second, we recorded the
frequencies of sampled recombinant edges, grouping them by the associated
donor and recipient branches in the clonal frame \citep{Didelot2010}, and
took ratios of these sampled counts with corresponding expected values
under the assumed 
prior distribution.  Third, we recorded the frequencies of these edges
types per site.  Finally, we defined a general, branch-independent {\em
  recombination intensity} at each site by summing the posterior
probabilities of all recombination edges.  In some cases, we also computed
intensities for particular types of replacing transfers, such as those
between the SPY and SDE clades, or those producing topologies different
from the clonal frame
(see \textbf{Supplementary Material}).

{\bf Significance testing}.  We used simulations to assess the significance
of the
prior-normalized counts of recombinant edges.  First,
we generated 100 replicate data sets under the prior model, with the same
block number and blocks lengths as the real data.  We then applied our
entire inference procedure to these data sets, and computed the same ratios
of posterior estimates to prior expectations as for the real data.  We then
compared the ratios estimated from real data with this empirical null
distribution.  Empirical one-sided $p$-values were computed as the fraction
of null estimates greater or less than the values estimated from real data.
This approach had the advantage of not only assessing the significance of
departures from prior expectations, but also correcting for any systematic
biases imposed by ClonalOrigin
(see \textbf{Supplementary Material}).

To compare the rates of SPY-to-SDE and SDE-to-SPY gene flow, we computed,
for each of the 1,001 genome-wide collections of recombinant trees, the
number of recombinant edges from any of the three SPY branches (SPY1, SPY2,
and SPY) to any of the three SDE branches (SDE1, SDE2, and SDE), and the
number of recombinant edges in the reverse direction.  We then took the
ratio of these two counts.  This gave us 1,001 samples from the posterior
distribution of the ratio of SPY-to-SDE and SDE-to-SPY recombinant edges,
from which we could estimate a posterior mean (2.0) and 95\% credible
interval (1.8--2.2).  None of the sampled values was as small as the prior
estimate of the ratio (1.4), indicating strong support in the data for a
preference for the SPY-to-SDE direction.

@@

{\bf Mowgli.} To study additive HGTs, we used a parsimony-based method
for reconciling gene and species trees called Mowgli \citep{Doyon2011}.
Given a gene tree and a species tree, Mowgli finds a reconciliation scenario 
that minimizes the number of gene duplication, loss, and
transfer (DLT) events, considering tree topologies with branching orders
(``labeled histories''; \citealp{Edwards1970}).  We estimated
unrooted gene trees for the nucleotide alignments for each of our
putative non-recombining gene fragment using RAxML \citep{Stamatakis2006},
then applied Mowgli to each of these trees together with the species
tree estimated by ClonalFrame (augmented with the SEE outgroup).
Since Mowgli required rooted gene trees, we ran Mowgli for all possible
rootings of each gene tree, and chose the scenario that produced the
minimum DLT score. In the case of multiple most parsimonious
reconciliations, one was chosen uniformly at random.

{\bf Distinguishing between replacing and additive HGTs.}  The transfer
events inferred by Mowgli had to be classified as replacing or additive in
a post-processing step because the program did not distinguish between
these types of events.  A naive approach would be to infer a replacing HGT
whenever an inferred transfer coincided exactly with an inferred loss in
the recipient lineage.  However, we found that loss events were difficult
to place accurately because they were frequently pushed toward the root of
the tree due to reconstruction error or parallel losses in multiple
lineages.  Therefore, we classified a transfer event as additive whenever a
descendant of the recipient lineage in the species tree contained both a
gene descended from the transferred copy and a gene not descended from that
copy.  All other transfers were considered replacing HGTs.  This approach
was conservative about calling additive HGTs, and labeled any HGT that
could plausibly be explained by replacing transfers as such (see
supplementary fig.\ \ref{fig:calling-transfers}).

{\bf Statistical enrichment}. As with replacing HGTs, we classified
transfer events by donor and recipient branch, and recorded the number of
inferred additive HGTs of each type.  We compared these numbers to
expectations based on simulations that assumed constant rates of
duplication, loss, and transfer across the species tree, given the total
numbers of events inferred across all 2,314 gene families divided by the
total branch lengths of the trees (see \textbf{Supplementary Material})
% (supplementary fig.\ \ref{fig:mowgli-sim}).

@@

We assigned genes to Gene Ontology (GO) categories by comparing the {\em
  Streptococcus} genes to bacterial proteins from the Uniref90 database
using BLASTP, and then assigning the same GO classification as the target
gene of the uniProt GOA database if the match had an $E$-value of
$<1.0\times10^{-5}.$  Virulence genes were identified using the methods
described by \cite{Suzuki2011}. A gene family was assigned a given
classification if any of its genes was assigned that classification.  To
test for associations with replacing gene transfers we used the
recombination intensity estimated for each gene, and for additive gene
transfers we used the number of additive transfers inferred for each
family.  To test for significance, we performed a Mann-Whitney $U$ test of
the values (recombination intensities or numbers of transfers) associated
with a given category vs.\ the values for the all other genes/families.  We
used the \cite{Benjamini1995} method to correct for multiple comparisons.

@@

\subsection*{Clonal frame and global parameter estimates}
% The 5-species tree was
% (((1:0.045557,2:0.045557)6:0.195330,3:0.240887)8:0.089874,(4:0.074544,5:0.074544)7:0.256218)9:0.000000
By applying ClonalFrame to our five-way genomic alignments
we obtained a phylogeny in which the two SPY
samples grouped together, as did the two SDE samples, and in which SDD and
SDE formed a clade, with SPY as an outgroup (fig.\ \ref{fig:tree5}).  This
species tree was consistent with previous results from 16S rRNA sequences
\citep{Facklam2002} and with general patterns observed genome-wide
\citep{Suzuki2011}.  The estimated levels of inter-species divergence were
substantial, at 0.48 substitution per site for SDD and SDE, and 0.66
substitutions per site for SPY and SDD/SDE.  An initial analysis with
ClonalOrigin (see \textbf{Methods}) produced an estimated per-site
population-scaled mutation rate of $\theta = $ 0.081 [interquartile range
across blocks: (0.067, 0.094)], a per-site population-scaled recombination
rate of $\rho = $ 0.012 (0.006, 0.019), and an average recombinant tract
length of $\delta = $ 744 (346, 2848) bp.  The full distributions of these
quantities across alignment blocks are shown in supplementary figure
\ref{fig:scatter3}.  A second run of the MCMC sampler produced nearly
identical results.  For comparison,
\cite{Didelot2010} obtained estimates of $\theta = 0.044$, $\rho = 0.017$,
and $\delta = 236$ for the {\em Bacillus cereus} group.  Our estimates
implied a ratio of $\rho/\theta = 0.15$, considerably lower than Didelot et
al.'s \citeyearpar{Didelot2010} estimate of $\rho/\theta = 0.405$ for {\em
  B.\ cereus}, suggesting increased rates of mutation and/or decreased
rates of recombination after controlling for differences in effective
population size.  We found that estimates of the average recombinant tract
length, $\delta$, were quite sensitive to our threshold for minimum block
length, with the inclusion of short alignments producing much larger
estimates due to edge effects.  We experimented with a range of thresholds
and selected a value (1500 bp) at which the estimates stabilized.

@@

There has been a great deal of interest for decades in the process of
horizontal gene transfer (HGT) and the manner in which it has influenced
phylogenetic relationships, particularly among bacteria \citep{Koonin2001}.
Most previous studies, however, have either simply examined patterns of
phylogenetic discordance, without differentiating between additive and
replacing HGTs \citep[e.g.,][]{Lerat2005}, or have focused specifically on
the process of homologous recombination, corresponding to replacing HGTs
\citep{Didelot2007,Didelot2010}.  To our knowledge, this is the first study
to examine both processes on a genome-wide scale, using both model- and
parsimony-based methods.
